Hilbert space methods to approximate Gaussian processes using Stan

### 2019-03-06 Paper/manuscript/manuscript_05.tex
more cleaning of the introduction

Gaussian processes (GPs) are flexible define probability distributions over multi-dimensional non-linear functions \citep{rasmussen2006gaussian,neal1997monte}. GPs are defind by a mean and covariance function and any finite set of function values are jointly distributed as a multivariate normal.  The covariance function encodes our prior assumptions about the functional relationship, such as continuity, smoothness, periodicity and scale properties. GPs not only allow for non-linear effects but can also implicitly handle interactions between covariates. Different types of covariance functions can be used, and they may be combined for further increased flexibility. Due to their generality and flexibility, GPs are of broad interest across machine learning and statistics \citep{rasmussen2006gaussian,neal1997monte}. Among others, they find application in the fields of spatial epidemiology \citep{diggle2013statistical,carlin2014hierarchical}, robotics and control \citep{deisenroth2015gaussian}, signal processing \citep{sarkka2013spatiotemporal}, as well as Bayesian optimization and probabilistic numerics \citep{roberts2010bayesian,briol2015probabilistic,hennig2015probabilistic}.

Given $n$ observations in the data, the computational cost and memory requirements of exact GP implementation in general scales as $O(n^3)$ and $O(n^2)$, respectively. This would limit their application to rather small data sets of a few tens of thousands observations at most. The problem becomes especially severe when performing full Bayesian inference via sampling methods, where in each sampling step we have $O(n^3)$ computational cost when inverting the Gram matrix of the covariance function, usually through Cholesky factorization. To alleviate these computational demands, several approximate methods have been proposed, which we will briefly review in the following.

Sparse GPs are based on low-rank approximations of the covariance matrix. The low-rank approximation with $m \ll n$ {\it inducing points} can be presented with reduced memory cost $O(nm)$ and the corresponding computational complexity is $O(nm^2)$.
% While computational complexity is reduced considerably, ensuring convergence to the exact GP model is not necessarily straightforward (Paul: this may need a citation).
A unifying view on sparse GPs based on approximate generative methods
is provided in \cite{quinonero2005unifying}, while a general review
can be found in \cite{rasmussen2006gaussian}.
% More recent developments in the context of sparse GPs include a structured kernel interpolation method \citep{wilson2015kernel}. More recently, \citep{bui2017unifying} proposed to use inducing points-based sparse approximation methods to perform GP approximations at inference time rather than at modeling time.

% (Paul: This whole paragraph is unclear to me. Are all of these approaches approximations via splines, or are the splines just special cases of a more general method? In any case, I think the paragraph needs substantial revision).
Alternative low-rank approximation is based on forming a basis function approximation with $m \ll$. The basis functions are presented explicitly, but could also be used to form a low rank covariance matrix approximation. Common basis function  approximations are based on the spectral analysis and series expansions of Gaussian processes \citep{loeve1977probability,trees1968detection,adler1981geometry,cramer2013stationary}.
Certain spline smoothing basis functions are equivalent to GPs with certain covariance function \citep{wahba1990spline}, but not all spline models are GPs (sometimes splines are not marginalization consistent)-
%
The sparse spectrum GPs are based on a sparse approximation to the frequency domain representation of a GP \citep{lazaro2010sparse,quia2010sparse,gal2015improving,gal2015improving}. Recently \citep{hensman2017variational} presented a variational Fourier feature approximation for Gaussian processes that was derived for the Mat{\'e}rn class of kernels. Random fourier features \citep{rahimi2008random,rahimi2009weighted} is a method for approximating kernels.

In this paper we focus on Laplace eigenfunction basis function approximation for stationary covariance functions proposed by \citet{solin2018hilbert}. Basis function approaches are computationally as linear models, which is attractive property in modular probabilistic programming models where there is a big benefit id approximation specific computation is simple. The Laplace eigenfunctions can be computed analytically and they are independent of the particular choice of the covariance kernel including the hyperparameters. While the pre-computation cost of the basis functions is $O(m^2n)$ the computation cost when only the covariance function parameters change is $O(mn+m)$ (or $O(m^3)$ if marginalization can be used). This is a big advantage for iterative algorithms such as Markov chain Monte Carlo (MCMC). The further advantage is the reduced memory cost in automatic differentation used in modern probabilistic programming frameworks, as the automatic differentation memory cost scales as the computational cost and not as the usual memory cost for the posterior density computation. The basis function approach provides also easy way to use non-centered parameterization reducing posterior dependency between parameters presenting the function and the covariance function parameters, which helps to further improve MCMC efficiency.

While \citet{solin2018hilbert} fully developed the mathematical theory behind the specific reduced-rank approximation, for practical implementations of approximate GPs in probabilistic programming framework further work is needed. In this paper we analyze in detail the performance and accuracy of the method in relation to key factors such as the number of basis functions, desired prediction space, or properties of the true functional relationship between covariates and response variable. We provide intuitive visualizations and practical recommendations for the choice of these factors, which will help users to improve computational performance while maintaining close approximation to exact GPs.

Although there are several GP specific software packages available to date (GPML, GPstuff, GPy, GPflow), each provide efficient implementations only for a restricted range of GP based models. In this paper we do not focus on the fastest possible inference for some specific GP models, but instead are interested in how GPs can be easily used as modular components in probabilistic programming frameworks such as Stan \citep{carpenter2017stan}.

@@

(Paul: The content and the purpose of the following sentences is not fully clear to me. Do we need them in the Introduction or can they be removed, or put into the next section? Gabi: In this paragraph I tried to state two differences I found in the literature between our method and the Sparse Spectrum GP and the Splines method. But I not know where locate them.) While Sparse Spectrum GP is based on a sparse spectrum, the reduced-rank method proposed in this paper aims to make the spectrum as ‘full’ as possible at a given rank. Recent Splines models can reproduce the Matern family of covariance functions (see, e.g., \cite{wood2003thin}), however our approach can reproduce basically all of the stationary covariance functions.

>>>>

Gaussian processes (GPs) are flexible statistical models for specifying probability distributions over multi-dimensional non-linear functions \citep{rasmussen2006gaussian,neal1997monte}. Their name stems from the fact that any finite set of function values is jointly distributed as a multivariate normal. GPs themselves are defined by two functions, a mean and a covariance function, with the latter being arguably the more important one. The covariance function encodes our prior assumptions about the functional relationship, such as continuity, smoothness, periodicity and scale properties. GPs not only allow for non-linear effects but can also implicitly handle interactions between covariates. Different types of covariance functions can be used, and they may be combined for further increased flexibility. Due to their generality and flexibility, GPs are of broad interest across machine learning and statistics \citep{rasmussen2006gaussian,neal1997monte}. Among others, they find application in the fields of spatial epidemiology \citep{diggle2013statistical,carlin2014hierarchical}, robotics and control \citep{deisenroth2015gaussian}, signal processing \citep{sarkka2013spatiotemporal}, as well as Bayesian optimization and probabilistic numerics \citep{roberts2010bayesian,briol2015probabilistic,hennig2015probabilistic}.

Given $n$ observations in the data, the computational comlexity and memory requirements of exact GP implementation in general scale as $O(n^3)$ and $O(n^2)$, respectively. This would limit their application to rather small data sets of a few tens of thousands observations at most. The problem becomes especially severe when performing full Bayesian inference via sampling methods, where in each sampling step we need $O(n^3)$ computations when inverting the Gram matrix of the covariance function, usually through Cholesky factorization. To alleviate these computational demands, several approximate methods have been proposed, which we will briefly review in the following.

Sparse GPs are based on low-rank approximations of the covariance matrix. The low-rank approximation with $m \ll n$ {\it inducing points} implies reduced memory requirements of $O(nm)$ and corresponding computational complexity of $O(nm^2)$.
A unifying view on sparse GPs based on approximate generative methods
is provided in \cite{quinonero2005unifying}, while a general review
can be found in \cite{rasmussen2006gaussian}.
% More recent developments in the context of sparse GPs include a structured kernel interpolation method \citep{wilson2015kernel}. More recently, \citep{bui2017unifying} proposed to use inducing points-based sparse approximation methods to perform GP approximations at inference time rather than at modeling time.

An alternative class of low-rank approximations is based on forming a basis function approximation with $m \ll n$ basis functions. The basis functions are usually presented explicitly, but can also be used to form a low rank covariance matrix approximation. Common basis function approximations rest on the spectral analysis and series expansions of Gaussian processes \citep{loeve1977probability,trees1968detection,adler1981geometry,cramer2013stationary}.
Sparse spectrum GPs are based on a sparse approximation to the frequency domain representation of a GP \citep{lazaro2010sparse,quia2010sparse,gal2015improving}. Recently, \cite{hensman2017variational} presented a variational Fourier feature approximation for Gaussian processes that was derived for the Mat{\'e}rn class of kernels. Another related method for approximating kernels relies on random fourier features \citep{rahimi2008random,rahimi2009weighted}.
Further, certain spline smoothing basis functions are equivalent to GPs with certain covariance functions \citep{wahba1990spline}, but not all spline models are GPs (e.g., sometimes splines are not marginalization consistent).

\textcolor{red}{Paul: Can we clarify how our discussed method relates to the former paragraph,
which introduces all sorts of basis function approximations. I.e. is our method
just one out of many or what makes it special? Perhaps we can somehow use the information provided
in: While Sparse Spectrum GP is based on a sparse spectrum, the reduced-rank method proposed in this paper aims to make the spectrum as ‘full’ as possible at a given rank. Recent Splines models can reproduce the Matern family of covariance functions (see, e.g., \cite{wood2003thin}), however our approach can reproduce basically all of the stationary covariance functions.}

In this paper we focus on the basis function approximation via Laplace eigenfunctions for stationary covariance functions proposed by \citet{solin2018hilbert}. Basis function approaches behave computationally like linear models, which is an attractive property in modular probabilistic programming models where there is a big benefit id approximation specific computation is simple (Paul: What do you mean with the the last part of the sentence?). The Laplace eigenfunctions can be computed analytically and they are independent of the particular choice of the covariance kernel including the hyperparameters. While the pre-computation cost of the basis functions is $O(m^2n)$, the computational cost when only changing the covariance function parameters is $O(mn+m)$ (or even $O(m^3)$ if marginalization can be used). This is a big advantage in terms of speed for iterative algorithms such as Markov chain Monte Carlo (MCMC). Another advantage is the reduced memory requirements of automatic differentation methods used in modern probabilistic programming frameworks (cite!). This is because the memory requirements of automatic differentation rather scale with the computational complexity instead of with the usual memory requirements for the posterior density computation. The discussed basis function approach also provides an easy way to apply the non-centered parameterization of GPs, which reduces the posterior dependency between parameters representing the estimated function and the hyperparameters of the covariance function, which helps to further improve MCMC efficiency.

While \citet{solin2018hilbert} have fully developed the mathematical theory behind this specific approximation of GPs, further work is needed for its practical implementation in probabilistic programming frameworks such as Stan \citep{carpenter2017stan}. In this paper, we analyze in detail the performance and accuracy of the method in relation to key factors such as the number of basis functions, desired prediction space, or properties of the true functional relationship between covariates and response variable. We provide intuitive visualizations and practical recommendations for the choice of these factors, which will help users to improve computational performance while maintaining close approximation to exact GPs.

Although there are several GP specific software packages available to date (GPML, GPstuff, GPy, GPflow, cite them!), each provide efficient implementations only for a restricted range of GP based models. In this paper, we do not focus on the fastest possible inference for some specific GP models, but instead are interested in how GPs can be easily used as modular components in probabilistic programming frameworks.

@@
